---
title: "SOC-GA 2332 Intro to Stats Lab 12"
date: "12/1/2023"
output:
  html_document:
    df_print: paged
    theme: paper
    highlight: textmate
    toc: true
  pdf_document: 
    toc: true
---


<style type="text/css">

body{ 

    font-size: 16px;
    line-height: 1.7em;
    <!-- text-align: justify; -->

}

blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 16px;
    border: solid 1px;
}

h1 { font-size: 32px; }

h2 { font-size: 24px; }

h3 { font-size: 20px; }

</style>

<br>

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

library(pacman)
p_load(tidyverse, data.table, stargazer, kableExtra, 
       gridExtra, MatchIt, PSweight, AER, haven,
       ggthemes, srvyr,mice)
```

## Part 0: Logistics

* Feedback to replication project part 1 released
* You have two options for part 2 (due Dec. 15)
  + A replication of the original KOB decomposition, with Table 1, 2, and Figure 2 (see instructions)
  + You can either manually code your KOB decomposition function, or use `R` packages, such as `oaxaca`. You should consider using the `twofold` rather than `threefold` option. 

* Three-fold decomposition:

$$
\begin{aligned}
\Delta \bar{Y} &= \bar{\boldsymbol{X}}_A'\hat{\beta}_A - \bar{\boldsymbol{X}}_B\hat{\beta}_B \\
\Delta \bar{Y} &= \underbrace{(\bar{\boldsymbol{X}}_A - \bar{\boldsymbol{X}}_B)'\hat{\beta}_B}_{\text{endowments}} +  \underbrace{\bar{\boldsymbol{X}}_B'(\hat{\beta}_A - \hat{\beta}_B)}_{\text{coefficients}} + \underbrace{(\bar{\boldsymbol{X}}_A - \bar{\boldsymbol{X}}_B)'(\hat{\beta}_A - \hat{\beta}_B)}_{\text{interaction}}
\end{aligned}
$$  

* Two-fold decomposition:

$$
\begin{aligned}
\Delta \bar{Y} &= \bar{\boldsymbol{X}}_A'\hat{\beta}_A - \bar{\boldsymbol{X}}_B\hat{\beta}_B \\
\Delta \bar{Y} &= \underbrace{(\bar{\boldsymbol{X}}_A - \bar{\boldsymbol{X}}_B)'\hat{\beta}_A}_{\text{explained}} +  \underbrace{\bar{\boldsymbol{X}}_B'(\hat{\beta}_A - \hat{\beta}_B)}_{\text{unexplained}}
\end{aligned}
$$ 

* The second option is to extend the analysis in some meaningful way
  + For example, you may ask whether the changing racial gap or gender gap is different for people in high-earning and low-earning occupations
  + Or you may ask whether the observed pattern is true for the White-Hispanic racial gap etc.
  + Please talk to me or Siwei on the ideas you have; we can assess its feasibility
  
* Assignment 4 will be released tomorrow. Due date will be Dec. 20 (11:59pm)

## Part 1: Using Weights in Descriptive Statistics and Regression (for IPUMS data)

* Weights are important in deriving an unbiased estimate for the population parameter, especially when the target parameter is heterogeneous across subgroups, and some subgroups are over- or under-sampled.

* We will use the IPUMS PERWT (person weight) variable to demonstrate how to use weights for descriptive statistics and regression models. 

* First, it is useful to review the PERWT variable description:  

  + PERWT indicates how many persons in the U.S. population are represented by a given person in an IPUMS sample. It is generally a good idea to use PERWT when conducting a person-level analysis of any IPUMS sample. The use of PERWT is optional when analyzing one of the "flat" or unweighted IPUMS samples. Flat IPUMS samples include the 1% samples from 1850-1930, all samples from 1960, 1970, and 1980, the 1% unweighted samples from 1990 and 2000, the 10% 2010 sample, and any of the full count 100% census datasets. PERWT must be used to obtain nationally representative statistics for person-level analyses of any sample other than those. 
  
* In `R`, there are multiple packages that can be used to generate weighted descriptive statistics such as `TAM`, `stats`, `survey` and `srvyr`. Here we use `srvyr` because it works well with `tidyverse` commands. For regression models, you can directly input the weights in the `lm()` function. 

```{r weights, warning=FALSE, message=FALSE}

## load data
wt_example <- read.csv("data/wt_example.csv")

## specify the survey design
wt_example <- wt_example %>%
  as_survey_design(weight = PERWT)

## summarization with and without weight
wt_example %>% 
  summarize(unweighted.mean = mean(INCWAGE),
            unweighted.sd = sd(INCWAGE),
            weighted.mean = survey_mean(INCWAGE),
            weighted.sd = survey_sd(INCWAGE)) %>%
    kbl("html") %>% 
  kable_classic_2(full_width = F)
```

* How is weighted mean calculated?

* The standard error of the weighted mean is actually a non-trivial issue
  + Check out [this site](https://www.alexstephenson.me/post/2022-04-02-weighted-variance-in-r/) for some brilliant discussions
  + People also use Taylor expansion to "linearize" the ratio form of the weighted mean and get standard error. Check out the section: Variance of the weighted mean in the [page](https://en.wikipedia.org/wiki/Weighted_arithmetic_mean#cite_note-sarndal1992-3)

```{r categorical weighted, warning=FALSE, message=FALSE}

## unweighted frequency table
wt_example %>%
  group_by(SEX) %>%
  summarize(freq = n()) %>%
  mutate(prop = freq/sum(freq)) %>%
    kbl("html") %>% 
  kable_classic_2(full_width = F)

## weighted frequency table
wt_example %>%
  group_by(SEX) %>%
  summarize(prop = survey_prop(),
            freq = survey_total()) %>%
    kbl("html") %>% 
  kable_classic_2(full_width = F)
```

```{r weighted regression}
## unweigthed regression
m1 <- lm(log(INCWAGE + 0.01) ~ SEX + EDUC + SEX*EDUC, data = wt_example)
## Weigthed regression
m2 <- lm(log(INCWAGE + 0.01) ~ SEX + EDUC + SEX*EDUC, data = wt_example, weights = PERWT)

## compare results
stargazer(m1, m2, 
          type = "text", 
          column.labels = c("Unweighted", "Weighted"), 
          dep.var.labels = "Log Wage")
```

* How is the weighted coefficient calculated?

## Part 2: Missing Data

* There are many ways to deal with missing data. The "correct" strategy will depend on your data, your research question and your methods. 

* But you should always be aware of the assumptions you make when inferring from a dataset with missing values

### 2.1 Complete-case analysis 

* The simplest way of dealing with missing data (and what most statistical software will do automatically) is to do **complete-case** analyses. In a complete-case analysis (listwise-deletion), you retain only those observations that are non-missing for all variables in your analysis. 
  + For example, the linear regression model in `R`, by default, uses complete cases
  + Pro: Simple, easy to execute 
  + Con: You may lose a significant amount of data that contain important information
  + It builds upon the very strong assumption of Missing Completely At Random (MCAR)

### 2.2 Mean and Conditional Mean (Regression) Imputation 

#### Mean Imputation 

* Apply the mean value of each variable to all missing cases of that variable

  + Mean imputation preserves the means of variables, but it assumes a more conservative distribution of the variables and tends to weaken relationships between variables
  + Mean imputation generally yields biased regression coefficients and invalid inferences
  + It ignores the possible relation between the missing variable and other variables in imputation
  + There are no estimates of uncertainties about the imputation

```{r impute mean, warning=FALSE}

## create some synthetic data with missingness
set.seed(87654)
N <- 1000
 
## some random variables
x1 <- round(rnorm(N), 2)
x2 <- round(x1 + rnorm(N, 10, 5))
x3 <- round(runif(N, -100, 20))
 
## manipulate missing values
x1[rbinom(N, 1, 0.2) == 1] <- NA  ## 20% missingness
x2[rbinom(N, 1, 0.05) == 1] <- NA ## 5% missingness
x3[rbinom(N, 1, 0.7) == 1] <- NA  ## 70% missingness
 
## indicator for missings (needed later)
x1_miss_ind <- is.na(x1)
x2_miss_ind <- is.na(x2)
x3_miss_ind <- is.na(x3)
 
## tore variables in a data frame
df_mean <- tibble(x1, x2, x3, x1_miss_ind, x2_miss_ind, x3_miss_ind)

## impute mean 
df_mean <- df_mean %>% 
  replace_na(list(x1 = mean(df_mean$x1, na.rm=T)))

## density of x1 pre and post imputation
 
### density of observed data
plot(density(df_mean$x1[x1_miss_ind == FALSE]),
     xlim = c(- 4, 4),
     ylim = c(0, 0.9),
     lwd = 2, 
     main = "Density Pre and Post Mean Imputation",
     xlab = "X1")
 
### density of observed & imputed data
points(density(df_mean$x1), 
       lwd = 2, 
       type = "l", 
       col = "red")
 
### legend
legend("topleft",
       c("Before Imputation", "After Imputation"),
       lty = 1,
       lwd = 2,
       col = c("black", "red"))
```

#### Conditional mean (regression) imputation

* Conditional-mean imputation replaces missing data with predicted values, obtained, for example, from a regression equation (regression imputation)
  + It provides the basis for Multiple Imputation, and frequently appears in the ML literature due to its "prediction" feature
  + In the simplest form, no uncertainties about imputation are introduced

### 2.3 Multiple Imputation with `mice`

```{r mice, warning=FALSE, message=FALSE}

## load data into the environment
data(wagepan, package = "wooldridge")

## take one-year sample 
dat <- wagepan %>%
  filter(year == 1980)  %>%
  select(c(agric, black, construc, ent, exper, fin, hisp, 
           poorhlth, married, hours, lwage))

## save the sample
original <- dat

## remove some values to create missingness
set.seed(1234)
dat[sample(1:nrow(dat), 20), "lwage"] <- NA

## initialize code
init <- mice(dat, maxit=0) 
meth <- init$method ## imputation method
predM <- init$predictorMatrix ## predictor matrix

## replace all except the variables we want to impute with ""
meth[!names(meth) == "lwage"] <- ""

## set imputation method
meth[c("lwage")] <- "norm"

## multiple imputation
set.seed(1234)
imputed_mice <- mice(dat, method = meth, predictorMatrix = predM, 
               m = 5, ## by default, m = 5
               maxit = 5) ## by default, maxit = 5 

## return the imputed data
imputed <- complete(imputed_mice) ## by default, the first imputed data is returned
imputed_complete <- complete(imputed_mice, "long") ## you can set `long` option to extract the complete m imputations

## check for missings
sum(is.na(imputed$lwage))

## check the predictions
actual <- original$lwage[is.na(dat$lwage)]
predicted <- imputed$lwage[is.na(dat$lwage)]
df <- tibble(actual, predicted)

## mean value
mean(actual)
mean(predicted)

## original and imputed
df

## regression analysis with pooled imputation
fit <- with(data = imputed_mice, 
            exp = lm(lwage ~ black + hisp + log(hours)))
summary(pool(fit))

## compare with the original regression
summary(lm(lwage ~ black + hisp + log(hours), dat))
```